---
title: "Alternative Software"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alternative Software}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- markdownlint-disable MD024 -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`rfastlowess` is presented throughout this package's documentation as a
faster, more feature-rich alternative to base R's `stats::lowess()`. This
page is for readers who already use `stats::lowess()` and want to know:
*can I get the exact same numbers out of `rfastlowess`, and if not, why
not?*

In short:

- With two options set (`boundary_policy = "noboundary"` and
  `scaling_method = "mar"`), `rfastlowess` reproduces `stats::lowess()` to
  within floating-point tolerance (~1e-10) — see
  [Reproducing `stats::lowess()`] below.
- Outside of those two options, `rfastlowess`'s *defaults* intentionally
  differ from `stats::lowess()` (see [Why the defaults differ]), and it
  supports a number of features `stats::lowess()` doesn't have (see
  [What this package adds]).

---

## Reproducing `stats::lowess()`

`stats::lowess()` and `rfastlowess`'s `Lowess()` implement the same
underlying algorithm, but two defaults differ:

| Option | `stats::lowess()` | `Lowess()` default |
| --- | --- | --- |
| Boundary padding | None (`"noboundary"`) | `"extend"` |
| Residual scaling | MAR: `median(\|r\|)` | `"mad"`: `median(\|r - median(r)\|)` |
| `fraction` | `f = 2/3` | `0.67` |
| `iterations` | `iter = 3` | `3` |
| `delta` | `0.01 * diff(range(x))` | same (`NULL` -> auto) |

Setting the first two options to match R, `Lowess()` reproduces
`stats::lowess()` exactly (up to floating-point rounding):

```{r alt_software_1}
library(rfastlowess)
set.seed(42)
x <- seq(0, 2 * pi, length.out = 60)
y <- sin(x) + rnorm(60, sd = 0.2)

reference <- stats::lowess(x, y, f = 2 / 3, iter = 3)

model <- Lowess(
    fraction = 2 / 3,
    iterations = 3L,
    boundary_policy = "noboundary",
    scaling_method = "mar"
)
result <- fit(model, x, y)

cat("Max abs difference:", max(abs(result$y - reference$y)), "\n")
```

If your `x` is not already sorted, add `outputs = "sorted"` to get results
ordered the same way `stats::lowess()` returns them (ascending by `x`):

```{r alt_software_2}
set.seed(1)
x <- seq(0, 2 * pi, length.out = 60)
x_unsorted <- sample(x)
y_unsorted <- sin(x_unsorted) + rnorm(60, sd = 0.2)

reference <- stats::lowess(x_unsorted, y_unsorted, f = 2 / 3, iter = 3)

model <- Lowess(
    fraction = 2 / 3,
    iterations = 3L,
    boundary_policy = "noboundary",
    scaling_method = "mar",
    outputs = "sorted"
)
result <- fit(model, x_unsorted, y_unsorted)

cat("x is sorted ascending:", all(diff(result$x) >= 0), "\n")
cat("Max abs difference:", max(abs(result$y - reference$y)), "\n")
```

Without `outputs = "sorted"`, `Lowess()` still fits on sorted `x` internally
(as `stats::lowess()` requires) but returns values reordered back to match
your original input order — useful when you need the fit aligned with other
columns in a data frame, but not what you want for a value-by-value
comparison against `stats::lowess()`'s own (always sorted) output.

---

## Why the defaults differ

`rfastlowess`'s *defaults* (as opposed to what it's capable of reproducing)
deliberately depart from `stats::lowess()` in two places:

**Boundary padding** (default `boundary_policy = "extend"` vs. no padding).
Without padding, the local neighbourhood at the first and last few points is
one-sided, which biases the fit toward the interior and increases variance
right at the edges. `"extend"` (and the other padding policies — see
`vignette("boundary")`) mitigate this at the cost of no longer being a
direct reproduction of Cleveland's original algorithm. `"noboundary"` is
kept as an explicit option specifically so reference-matching remains
possible.

**Residual scaling** (default `scaling_method = "mad"` vs. MAR). MAD
(`median(|r - median(r)|)`) centers residuals at their median before taking
the median absolute value, which is a breakdown-point-optimal scale
estimator; MAR (`median(|r|)`, what R uses) does not center first, so it can
be biased when residuals are systematically skewed. See
`vignette("scaling")` for the full comparison, including `"mean"`.

---

## What this package adds

`stats::lowess()` doesn't support:

| Feature | `rfastlowess` | `stats::lowess()` |
| --- | :---: | :---: |
| Kernel functions | 7 options | Tricube only |
| Robustness weighting | 3 options | Bisquare only |
| Scale estimation | MAD, MAR, mean | MAR only |
| Boundary padding | 4 policies | none |
| Confidence / prediction intervals | yes | no |
| Cross-validation for `fraction` | K-fold, LOOCV | no |
| Streaming / online modes | yes | no |
| Custom per-observation weights | yes | no |
| Parallel / GPU execution | yes | no |

See `vignette("concepts")` for an overview of these, or the
[Benchmarks](benchmarks.html) article for performance comparisons.
