---
title: "Benchmarks"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## CPU Benchmarks

Speedup relative to R's `stats::lowess` (higher is better):

| Category | R baseline | rfastlowess Serial | rfastlowess Parallel |
| --- | --- | --- | --- |
| **Clustered** | 2.34 ms | 2.0× | **2.5×** |
| **Constant Y** | 1.81 ms | 1.7× | **3.2×** |
| **Extreme Outliers** | 5.81 ms | 1.5× | **2.6×** |
| **Financial** (500–5K) | 0.65 ms | **2.0×** | 1.4× |
| **Fraction** (0.05–0.67) | 3.8 ms | 1.6× | **3.2×** |
| **Genomic** (1K–100K) | 11.2 ms | 2.2× | **2.4×** |
| **High Noise** | 7.08 ms | 1.5× | **3.6×** |
| **Iterations** (0–10) | 3.0 ms | 1.9× | **2.7×** |
| **Scale** (1K–10K) | 1.6 ms | 1.5× | **1.6×** |
| **Scientific** (500–5K) | 0.9 ms | 1.4× | 1.4× |

*The R column shows the average time across scenarios in multi-scenario categories. Speedups are averages across the same range.*

---

## GPU Backend

For large batch datasets, the GPU backend can outperform CPU-parallel execution. The crossover point depends on `fraction × n`:

| Scenario | CPU-Parallel | GPU | Speedup |
| --- | --- | --- | --- |
| n = 1M, fraction = 0.5 | 1.24 s | 187 ms | **6.6×** |

At `fraction = 0.5`, GPU overtakes CPU around n ≥ 50K; at smaller fractions, around n ≥ 100K–250K. See [benchmarks/README.md](https://github.com/thisisamirv/lowess-project/blob/main/benchmarks/README.md#gpu-benchmarks) for the full sweep and transfer-overhead breakdown.

See the [benchmarks README](https://github.com/thisisamirv/lowess-project/blob/main/benchmarks/README.md) for full sweep details and transfer-overhead breakdown.

---

## Reproducing Benchmarks

The benchmarks compare `rfastlowess` against `stats::lowess` using the `microbenchmark` package.

```{r benchmarks_1, eval=FALSE}
# Install benchmark dependency
# install.packages("microbenchmark")

library(rfastlowess)
library(microbenchmark)

set.seed(42)
n <- 5000
x <- seq(0, 10, length.out = n)
y <- sin(x) + rnorm(n, sd = 0.3)

mb <- microbenchmark(
    stats_lowess = stats::lowess(x, y, f = 0.67),
    rfastlowess_serial = {
        m <- Lowess(fraction = 0.67)
        fit(m, x, y)
    },
    rfastlowess_parallel = {
        m <- Lowess(fraction = 0.67, parallel = TRUE)
        fit(m, x, y)
    },
    times = 50
)
print(mb)
```

```{r session-info}
sessionInfo()
```
