---
title: "Boundary Handling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Boundary Handling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

![Boundary policy comparison](../man/figures/boundary_comparison.svg)

Standard LOWESS neighbourhoods become asymmetric at the boundaries: fewer points
exist on one side, pulling the local fit toward the data interior. The
`boundary_policy` parameter controls how the data is padded to mitigate this
effect.

| Policy | Padding Strategy | Best For |
| --- | --- | --- |
| `"extend"` | Repeat first / last value | Most datasets (default) |
| `"reflect"` | Mirror data at boundaries | Periodic or symmetric data |
| `"zero"` | Pad with zeros | Data known to approach zero |
| `"noboundary"` | No padding (Cleveland original) | Reference behaviour |

---

## Extend (Default)

Pads beyond both endpoints by replicating the first and last observed values.
Prevents the fit from curling toward zero.

**Use when**: No strong prior on boundary behaviour; general-purpose smoothing.

```{r boundary_1}
library(rfastlowess)
set.seed(42)
x <- seq(0, 2 * pi, length.out = 100)
y <- sin(x) + rnorm(100, sd = 0.3)

model <- Lowess(boundary_policy = "extend")
result <- fit(model, x, y)
cat("First 6 smoothed values (extend policy):\n")
print(head(result$y))
```

---

## Reflect

Mirrors data at both endpoints. Prevents inflection artifacts when the signal is
periodic or symmetric.

**Use when**: Signal is periodic (e.g. time-of-day) or the boundary is a known
symmetry point.

```{r boundary_2}
model <- Lowess(boundary_policy = "reflect")
result <- fit(model, x, y)
cat("First 6 smoothed values (reflect policy):\n")
print(head(result$y))
```

---

## Zero

Pads with zeros beyond the endpoints.

**Use when**: The signal is known to decay to zero at the boundaries (e.g.
impulse responses, genomic signals at chromosome ends).

```{r boundary_3}
model <- Lowess(boundary_policy = "zero")
result <- fit(model, x, y)
cat("First 6 smoothed values (zero policy):\n")
print(head(result$y))
```

---

## No Boundary Padding

Reproduces Cleveland's original behaviour — no padding applied.

**Use when**: Comparing against reference implementations; the original
Cleveland algorithm behaviour is required.

> **Note:** Without padding, boundary fits can have higher variance and visible
> edge artefacts, particularly with small `fraction` values.

```{r boundary_4}
model <- Lowess(boundary_policy = "noboundary")
result <- fit(model, x, y)
cat("First 6 smoothed values (noboundary policy):\n")
print(head(result$y))
```

---

## Comparing Policies

```{r boundary_5}
library(rfastlowess)
set.seed(42)
x <- seq(0, 2 * pi, length.out = 100)
y <- sin(x) + rnorm(100, sd = 0.3)

policies <- c("extend", "reflect", "zero", "noboundary")
colors   <- c("blue", "red", "green", "purple")

plot(x, y, pch = 16, col = "gray",
    main = "Boundary Policy Comparison")

for (i in seq_along(policies)) {
    model  <- Lowess(boundary_policy = policies[i])
    result <- fit(model, x, y)
    lines(result$x, result$y, col = colors[i], lwd = 2)
}

legend("topright", policies, col = colors, lwd = 2)
```

```{r session-info}
sessionInfo()
```
