---
title: "LOWESS Concepts"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{LOWESS Concepts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## What is LOWESS?

**LOWESS** (Locally Weighted Scatterplot Smoothing) is a nonparametric regression method that fits smooth curves through scatter plots without assuming a global functional form.

Unlike parametric methods (linear regression, polynomial fitting), LOWESS adapts locally to the data structure, making it ideal for:

- **Exploratory data analysis** — Discover patterns without assumptions
- **Trend estimation** — Extract signals from noisy time series
- **Baseline correction** — Remove systematic effects in spectroscopy
- **Genomic smoothing** — Smooth methylation, ChIP-seq, or expression data

---

## How It Works

![LOWESS smoothing concept](../man/figures/lowess_concept.svg)

For each point in your data, LOWESS:

1. **Selects neighbors** — Choose the nearest points (controlled by `fraction`)
2. **Assigns weights** — Closer points get higher weights (using a kernel function)
3. **Fits locally** — Perform weighted least squares regression
4. **Extracts value** — Use the fitted value as the smoothed estimate
5. **Iterates** (optional) — Reweight points based on residuals to reduce outlier influence

---

## The Fraction Parameter

The `fraction` (also called bandwidth or span) is the most important parameter. It controls what proportion of data is used for each local fit.

![Effect of the fraction parameter](../man/figures/fraction_comparison.svg)

| Fraction | Effect | When to Use |
| --- | --- | --- |
| **0.1–0.3** | Fine detail, follows data closely | Rapidly changing signals |
| **0.3–0.5** | Balanced smoothing | Most applications |
| **0.5–0.7** | Heavy smoothing | Noisy data, trend extraction |
| **0.7–1.0** | Very smooth | Strong noise, global trends |

> **Tip:** Start with `fraction=0.67` (the default) and adjust based on visual inspection. Use cross-validation for automated selection.

---

## Robustness Iterations

Standard LOWESS is sensitive to outliers. **Robustness iterations** downweight points with large residuals:

![Effect of robustness iterations](../man/figures/robust_iter_comparison.svg)

| Iterations | Effect | When to Use |
| --- | --- | --- |
| **0** | No robustness (fastest) | Clean data, speed-critical |
| **1–3** | Moderate robustness | Most applications |
| **4–6** | Strong robustness | Data with outliers |
| **7+** | Very strong | Heavy contamination |

---

## Confidence vs Prediction Intervals

![Confidence and prediction intervals](../man/figures/intervals_comparison.svg)

| Interval Type | What It Represents | Width |
| --- | --- | --- |
| **Confidence** | Uncertainty in the *mean curve* | Narrow |
| **Prediction** | Uncertainty for *new observations* | Wide |

- Use **confidence intervals** to show where the true trend likely lies
- Use **prediction intervals** to show where new data points might fall

---

## Execution Modes

Choose the right mode based on your use case:

| Mode | Use Case | Memory | Features |
| --- | --- | --- | --- |
| **Batch** | Complete datasets | Entire dataset | All features |
| **Streaming** | Large files (>100K points) | One chunk | Residuals, robustness |
| **Online** | Real-time data | Fixed window | Incremental updates |

Decision guide:

- Data fits in memory and no real-time requirement → **Batch**
- Data does not fit in memory → **Streaming**
- Real-time / point-by-point processing needed → **Online**

---

## Key Advantages

| Feature | LOWESS | Polynomial Regression | Moving Average |
| --- | --- | --- | --- |
| No parametric assumptions | ✓ | ✗ | ✓ |
| Adapts to local structure | ✓ | ✗ | Partial |
| Robust to outliers | ✓ | ✗ | ✗ |
| Uncertainty estimates | ✓ | ✓ | ✗ |
| Handles irregular sampling | ✓ | ✓ | ✗ |

```{r session-info}
sessionInfo()
```
