---
title: "Cross-Validation"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Cross-Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

![Cross-validation comparison](../man/figures/cv_comparison.svg)

Cross-validation helps select the optimal `fraction` parameter by evaluating
performance on held-out data.

---

## K-Fold Cross-Validation

Split data into K folds, train on K-1, validate on 1. Repeat for each fold and
average the scores.

```{r cross_validation_1}
library(rfastlowess)
set.seed(42)
x <- seq(0, 2 * pi, length.out = 100)
y <- sin(x) + rnorm(100, sd = 0.3)

model <- Lowess(
    cv_method = "kfold",
    cv_k = 5,
    cv_fractions = c(0.2, 0.3, 0.5, 0.7)
)
result <- fit(model, x, y)

cat("Selected fraction:", result$fraction_used, "\n")
cat("CV scores:", result$cv_scores, "\n")
```

### Parameters

| Parameter | Default | Description |
| --- | --- | --- |
| `cv_method` | `NULL` | `"kfold"` or `"loocv"` |
| `cv_k` | 5 | Number of folds (k-fold only) |
| `cv_fractions` | automatic | Candidate fraction values to search |

---

## Leave-One-Out Cross-Validation (LOOCV)

Each point is held out once. More thorough but slower than k-fold.

```{r cross_validation_2}
library(rfastlowess)
set.seed(42)
x <- seq(0, 2 * pi, length.out = 100)
y <- sin(x) + rnorm(100, sd = 0.3)

model <- Lowess(
    cv_method = "loocv",
    cv_fractions = c(0.2, 0.3, 0.5, 0.7)
)
result <- fit(model, x, y)

cat("Selected fraction:", result$fraction_used, "\n")
```

---

## Custom Fraction Grid

Provide a fine grid of candidate fractions for more precise selection.

```{r cross_validation_3}
library(rfastlowess)
set.seed(42)
x <- seq(0, 2 * pi, length.out = 100)
y <- sin(x) + rnorm(100, sd = 0.3)

fractions <- seq(0.1, 0.9, by = 0.05)

model <- Lowess(
    cv_method = "kfold",
    cv_k = 10,
    cv_fractions = fractions
)
result <- fit(model, x, y)

cat("Best fraction:", result$fraction_used, "\n")
cat("CV scores:\n")
print(data.frame(fraction = fractions, score = result$cv_scores))
```

---

## Inspecting CV Results

After fitting, `result$cv_scores` contains the CV score for each candidate
fraction in the order they were passed.

```{r cross_validation_4}
library(rfastlowess)
set.seed(42)
x <- seq(0, 2 * pi, length.out = 100)
y <- sin(x) + rnorm(100, sd = 0.3)

fractions <- c(0.1, 0.2, 0.3, 0.5, 0.7)
model <- Lowess(cv_method = "kfold", cv_fractions = fractions)
result <- fit(model, x, y)

plot(fractions, result$cv_scores, type = "b",
     xlab = "Fraction", ylab = "CV Score (MSE)",
     main = "Cross-Validation Score by Fraction")
abline(v = result$fraction_used, col = "red", lty = 2)
```

---

## Seeded Randomization

Set a seed for reproducible fold assignments in k-fold CV:

```{r cross_validation_5}
library(rfastlowess)
set.seed(42)
x <- seq(0, 2 * pi, length.out = 100)
y <- sin(x) + rnorm(100, sd = 0.3)

model <- Lowess(
    cv_method = "kfold",
    cv_k = 5,
    cv_fractions = c(0.2, 0.3, 0.5, 0.7),
    cv_seed = 42L
)
result <- fit(model, x, y)
cat("Selected fraction:", result$fraction_used, "\n")
```

---

## Method Comparison

| Method | Folds | Speed | Variance | Bias |
| --- | --- | --- | --- | --- |
| **KFold(5)** | 5 | Fast | Moderate | Low |
| **KFold(10)** | 10 | Medium | Lower | Lower |
| **LOOCV** | N | Slow | Lowest | Lowest |

Use **5-fold** or **10-fold** CV for most applications. LOOCV is only worth it
for small datasets (N < 100).

```{r session-info}
sessionInfo()
```
