---
title: "Custom Weights"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Weights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- markdownlint-disable MD024 -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## How Custom Weights Work

Standard LOWESS assigns equal prior trust to all observations. Custom weights
let you override this assumption point by point — before any distance or
robustness weighting is applied.

The effective weight of observation $j$ in a local fit centred at $x_i$ is:

$$w_{ij} = \text{custom\_weights}[j] \times K\!\left(\frac{d_{ij}}{h_i}\right)
\times r_j$$

where $K$ is the distance kernel, $h_i$ is the local bandwidth, and $r_j$ is the
robustness weight from the current iteration.

> **Batch adapter only:** `custom_weights` applies in **Batch** mode. It is
> silently ignored in Streaming and Online adapters.

---

## When to Use Custom Weights

| Situation | Recommended weight |
| --- | --- |
| Point known to be erroneous | `0.0` — fully excluded |
| Unreliable sensor / low precision | `0.1 – 0.5` |
| Standard observation | `1.0` (default) |
| Carefully calibrated measurement | `> 1.0` |
| Measurement uncertainty $\sigma_i$ | $1 / \sigma_i^2$ |

### Custom Weights vs. Robustness Iterations

Both mechanisms handle unreliable data, but they serve different purposes:

| | Custom Weights | Robustness Iterations |
| --- | --- | --- |
| **When known** | Before fitting | Computed from residuals |
| **Knowledge required** | Prior knowledge of quality | None — data-driven |
| **Effect** | Fixed throughout fit | Adapts each iteration |
| **Use case** | Known bad sensors | Unknown outlier contamination |

They compose: you can use both simultaneously. Custom weights suppress
*a priori* bad points; robustness iterations then handle any *residual*
outliers that remain.

---

## Basic Usage

### Suppress a Known Outlier

Set the weight to `0` at the bad point — it is excluded from every local fit
that would otherwise include it.

```{r custom_weights_1}
library(rfastlowess)

x <- 1:10
y <- x * 2.0
y[6] <- 100.0              # spike at index 6

weights <- rep(1.0, 10)
weights[6] <- 0.0          # exclude the spike

model <- Lowess(fraction = 0.5, iterations = 0L)
result <- fit(model, x, y, custom_weights = weights)
cat("Smoothed y[0]:", result$y[1], "\n")
```

---

### Emphasize Important Points

Assign high weights to measurements you trust most — calibration standards,
reference instruments, or low-noise observations.

```{r custom_weights_2}
library(rfastlowess)
x <- seq(0, 2 * pi, length.out = 100)
y <- sin(x) + 0.1
calibration_indices <- c(6, 21, 41, 61, 81)

weights <- rep(1.0, length(x))
weights[calibration_indices] <- 10.0

model <- Lowess(fraction = 0.5)
result <- fit(model, x, y, custom_weights = weights)
cat("Smoothed y[0]:", result$y[1], "\n")
```

---

### Propagate Measurement Uncertainty

If each observation has a known standard deviation $\sigma_i$, set
$w_i = 1 / \sigma_i^2$ to give the fit information-theoretically optimal
weighting.

```{r custom_weights_3}
library(rfastlowess)
x <- seq(0, 2 * pi, length.out = 100)
y <- sin(x) + 0.1
sigma <- 0.1 + (seq_along(x) %% 4) * 0.1

weights <- 1 / sigma^2
model <- Lowess(fraction = 0.5)
result <- fit(model, x, y, custom_weights = weights)

plot(x, y, pch = 16, cex = 0.5 + weights / max(weights),
    col = "gray", main = "Inverse-Variance Weighted LOWESS")
lines(result$x, result$y, col = "blue", lwd = 2)

cat("Smoothed y[0]:", result$y[1], "\n")
```

---

## Combined with Robustness Iterations

Custom weights and robustness iterations compose naturally: use custom weights
for *known* bad points and robustness for *unknown* contamination.

```{r custom_weights_4}
library(rfastlowess)
x <- 0:19
y <- x * 1.5
y[4]  <- -50.0   # known bad
y[13] <- 80.0    # unknown outlier

weights <- rep(1.0, 20)
weights[4] <- 0.0

model <- Lowess(fraction = 0.4, iterations = 3)
result <- fit(model, x, y, custom_weights = weights)
cat("Smoothed y[0]:", result$y[1], "\n")
```

---

## Validation Rules

| Rule | Effect |
| --- | --- |
| Length must equal `n` | Error at fit time if mismatched |
| All values must be ≥ 0 | Negative weights are rejected |
| All-zero weight vector | Error: no points remain for any local fit |
| Uniform weights (`1.0` everywhere) | Identical result to omitting weights |

> **Zero-weight windows:** If a local neighbourhood contains only zero-weight
> points, the fit at that centre point falls back to the behaviour specified by
> `zero_weight_fallback` (default: `"use_local_mean"`).

---

## See Also

- [Robustness](robustness.html) — adaptive outlier downweighting via IRLS
- `?Lowess` — full parameter reference

```{r session-info}
sessionInfo()
```
