---
title: "Custom Weights"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Custom Weights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## How Custom Weights Work

Per-observation weights encode data quality directly into the LOWESS fit.
Standard LOWESS assigns equal prior trust to all observations. Custom weights
let you override this assumption point by point — before any distance or
robustness weighting is applied.

The effective weight of observation $j$ in a local fit centred at $x_i$ is:

$$w_{ij} = \text{custom\_weights}[j] \times K\!\left(\frac{d_{ij}}{h_i}\right)
\times r_j$$

where $K$ is the distance kernel, $h_i$ is the local bandwidth, and $r_j$ is the
robustness weight from the current iteration.

> **Note:** `custom_weights` applies in **Batch** mode only. It is silently
> ignored in Streaming and Online adapters.

---

## When to Use Custom Weights

| Situation | Recommended weight |
| --- | --- |
| Point known to be erroneous | `0.0` — fully excluded |
| Unreliable sensor / low precision | `0.1 – 0.5` |
| Standard observation | `1.0` (default) |
| Carefully calibrated measurement | `> 1.0` |
| Measurement uncertainty $\sigma_i$ | $1 / \sigma_i^2$ |

### Custom Weights vs. Robustness Iterations

| | Custom Weights | Robustness Iterations |
| --- | --- | --- |
| **When known** | Before fitting | Computed from residuals |
| **Knowledge required** | Prior knowledge of quality | None — data-driven |
| **Effect** | Fixed throughout fit | Adapts each iteration |
| **Use case** | Bad sensors, calibration | Unknown outlier contamination |

They compose: use both simultaneously. Custom weights suppress known bad points;
robustness iterations handle residual outliers.

---

## Suppress a Known Outlier

Set the weight to `0` at the bad point — it is excluded from every local fit.

```{r custom_weights_1}
library(rfastlowess)

x <- 1:10
y <- x * 2.0
y[6] <- 100.0              # spike at index 6

weights <- rep(1.0, 10)
weights[6] <- 0.0          # exclude the spike

model <- Lowess(fraction = 0.5, iterations = 0L)
result <- fit(model, x, y, custom_weights = weights)
print(head(result$y))
```

---

## Inverse-Variance Weights

Weight by $1/\sigma_i^2$ when measurement uncertainties are known.

```{r custom_weights_2}
library(rfastlowess)
set.seed(42)
x <- seq(0, 10, length.out = 100)
y <- sin(x) + rnorm(100, sd = 0.5)

# Measurement precision varies along x
sigma <- 0.1 + 0.5 * abs(sin(x))
weights <- 1 / sigma^2

model <- Lowess(fraction = 0.3)
result <- fit(model, x, y, custom_weights = weights)

plot(x, y, pch = 16, cex = 0.5 + weights / max(weights),
     col = "gray", main = "Inverse-Variance Weighted LOWESS")
lines(result$x, result$y, col = "blue", lwd = 2)
```

---

## Combining Custom Weights with Robustness

```{r custom_weights_3}
library(rfastlowess)
set.seed(42)
x <- 1:100
y <- sin(x / 10) + rnorm(100, sd = 0.3)

# Known bad region: indices 40–50
weights <- rep(1.0, 100)
weights[40:50] <- 0.1

# Also use robustness for unknown outliers
model <- Lowess(fraction = 0.3, iterations = 3)
result <- fit(model, x, y, custom_weights = weights)
print(head(result$y))
```

```{r session-info}
sessionInfo()
```
