---
title: "GPU Backend"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{GPU Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

The batch `Lowess` type can execute on a GPU-accelerated backend powered by `wgpu` (Vulkan / Metal / DX12). The GPU backend reimplements almost the entire LOWESS pipeline — local regression fitting, robustness iterations, interval bounds, and cross-validation — as compute shaders, so all anchor points are fitted in parallel.

> **Batch only.** GPU support applies to the batch `Lowess` type only.
> `StreamingLowess` and `OnlineLowess` remain CPU-only.

GPU acceleration is worth enabling for roughly 10K+ points. For smaller inputs the CPU backend (optionally with `parallel = TRUE`) is typically faster once GPU dispatch overhead is accounted for.

---

## Supported Features on GPU

| Feature | CPU | GPU |
| --- | --- | --- |
| Batch fitting | ✅ | ✅ |
| Streaming / Online | ✅ | ❌ |
| All weight / robustness / scaling methods | ✅ | ✅ |
| Confidence / prediction intervals | ✅ | ✅ |
| Cross-validation (k-fold, LOOCV) | ✅ | ✅ |
| Custom per-observation weights | ✅ | ✅ |

---

## Checking GPU Availability

```{r gpu_backend_1}
library(rfastlowess)

gpu_available()
```

Returns `TRUE` if the currently loaded library was built with GPU support, `FALSE` otherwise.

---

## Installing GPU Support

GPU support is opt-in and not included in standard CRAN / R-universe builds. Install it with a one-time download:

```{r gpu_backend_2, eval=FALSE}
library(rfastlowess)

install_gpu()          # interactive — prompts for confirmation
install_gpu(yes = TRUE) # non-interactive
```

`install_gpu()` downloads a prebuilt GPU-enabled library from the matching GitHub release and copies it into the installed package's `libs/` directory.

**Restart R after installation.**

Alternatively, build from source with GPU support:

```sh
make -f bindings/r/Makefile WITH_GPU=1
```

---

## Using the GPU Backend

Once GPU support is installed, pass `backend = "gpu"` to `Lowess()`:

```{r gpu_backend_3, eval=FALSE}
library(rfastlowess)
set.seed(42)
n <- 100000
x <- seq(0, 2 * pi, length.out = n)
y <- sin(x) + rnorm(n, sd = 0.3)

model <- Lowess(
    fraction  = 0.5,
    iterations = 3,
    backend   = "gpu"
)
result <- fit(model, x, y)
```

---

## Performance Guidelines

| Scenario | Recommendation |
| --- | --- |
| n < 10K | CPU (`parallel = TRUE`) |
| n 10K–100K | Try both; GPU wins for larger fractions |
| n > 100K | GPU backend recommended |
| `fraction = 0.5`, n ≥ 50K | GPU outperforms CPU-parallel |
| `fraction = 0.1`, n ≥ 250K | GPU crossover point |

At n = 1M (`fraction = 0.5`, 3 iterations), the GPU backend is approximately **6.6×** faster than CPU-parallel.

---

## Fallback Behaviour

If `backend = "gpu"` is requested but GPU support is not installed, an informative error is raised pointing to `install_gpu()`. No silent fallback to CPU occurs.

```{r session-info}
sessionInfo()
```
