---
title: "Merge Strategies"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Merge Strategies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- markdownlint-disable MD024 -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

Streaming LOWESS processes data in fixed-size chunks with a configurable overlap. Points inside the overlap zone are fitted twice — once by the left chunk and once by the right chunk. The `merge_strategy` decides how those two estimates are combined.

```text
Chunk A:   [=========|=====]
Chunk B:            [=====|=========]
Overlap:            [=====]
                      ↑
                 merge_strategy
                 applied here
```

| Strategy | Method | Robustness | Speed |
| --- | --- | --- | --- |
| `"average"` | Simple mean of both estimates | Low | Fastest |
| `"take_first"` | Left-chunk estimate only | Low | Fastest |
| `"take_last"` | Right-chunk estimate only | Low | Fastest |
| `"weighted_average"` | Distance-weighted mean (default) | High | Moderate |

![Merge strategy comparison](../man/figures/merge_comparison.svg)

---

## Weighted Average (Default)

Blends left- and right-chunk estimates using distance-based weights. Points closer to the centre of each chunk get higher weight.

**Use when**: General-purpose streaming; most reliable option.

```{r}
library(rfastlowess)
set.seed(42)
x <- seq(0, 2 * pi, length.out = 100)
y <- sin(x) + rnorm(100, sd = 0.3)
x_chunk <- x[seq_len(50)]
y_chunk <- y[seq_len(50)]

model <- StreamingLowess(
    merge_strategy = "weighted_average",
    chunk_size = 5000,
    overlap = 500
)
result <- process_chunk(model, x_chunk, y_chunk)
```

---

## Average

Takes the arithmetic mean of the left-chunk and right-chunk estimates in the overlap region.

**Use when**: Chunks are large and the overlap region has uniform data density.

```{r}
model <- StreamingLowess(
    merge_strategy = "average",
    chunk_size = 5000,
    overlap = 500
)
result <- process_chunk(model, x_chunk, y_chunk)
```

---

## Take First

Uses only the left-chunk estimate in the overlap region.

**Use when**: The left chunk has a larger context window and is expected to be more accurate at the overlap boundary.

```{r}
model <- StreamingLowess(
    merge_strategy = "take_first",
    chunk_size = 5000,
    overlap = 500
)
result <- process_chunk(model, x_chunk, y_chunk)
```

---

## Take Last

Uses only the right-chunk estimate in the overlap region.

**Use when**: The right chunk is expected to provide better context (e.g., more data follows).

```{r}
model <- StreamingLowess(
    merge_strategy = "take_last",
    chunk_size = 5000,
    overlap = 500
)
result <- process_chunk(model, x_chunk, y_chunk)
```

---

## Choosing Chunk Size and Overlap

| Consideration | Recommendation |
| --- | --- |
| Memory limit | `chunk_size` = largest chunk that fits comfortably |
| Edge accuracy | `overlap` ≥ `fraction × chunk_size` |
| Speed | Larger chunks → fewer merges → faster |

```{r}
# Rule of thumb: overlap = fraction * chunk_size
fraction  <- 0.3
chunk_size <- 5000
overlap    <- ceiling(fraction * chunk_size)  # 1500

model <- StreamingLowess(
    fraction       = fraction,
    chunk_size     = chunk_size,
    overlap        = overlap,
    merge_strategy = "weighted_average"
)
```

---

## Choosing a Strategy

| Situation | Recommended Strategy |
| --- | --- |
| General purpose | `"weighted_average"` |
| Maximum throughput | `"average"` |
| Immediate finalised output | `"take_first"` |
| Post-processing, right context better | `"take_last"` |
| Minimising boundary artefacts | `"weighted_average"` |
