---
title: "Online Mode (OnlineLowess)"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Online Mode (OnlineLowess)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **See also:** [Choosing an Execution Mode](adapter-choice.html) for a comparison of all three modes.

Maintains a sliding window and processes each incoming point immediately.

![Online adapter comparison](../man/figures/online_comparison.svg)

## When to Use

- Real-time data streams (sensors, logs)
- Each point must be smoothed as it arrives
- Memory-bounded processing with a fixed window

## Parameters

| Parameter | Default | Description |
| --- | --- | --- |
| `window_capacity` | 1000 | Max points in sliding window |
| `min_points` | 2 | Minimum points before output |
| `update_mode` | `"incremental"` | `"incremental"` or `"full"` refit |

## Example

```{r}
library(rfastlowess)
set.seed(42)
times <- 1:100
temperatures <- 20 + 5 * sin(times / 10) + rnorm(100)

model <- OnlineLowess(
    fraction = 0.3,
    window_capacity = 25,
    min_points = 5,
    update_mode = "incremental"
)
for (i in seq_along(times)) {
    result <- add_point(model, times[i], temperatures[i])
    if (!is.null(result))
        cat(sprintf("Time %d: %.2f\n", times[i], result$y))
}
```

> **Note:** `update_mode = "incremental"` refits only the most recent point
> for lower latency. `update_mode = "full"` refits the entire window for
> higher accuracy.
