---
title: "Parameters Reference"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Parameters Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## All Parameters

| Parameter | Default | Range | Description | Mode |
| --- | --- | --- | --- | --- |
| **fraction** | 0.67 | (0, 1] | Neighbourhood size | All |
| **iterations** | 3 | [0, 1000] | Robustness iterations | All |
| **delta** | `NULL` | [0, ∞) | Distance threshold for skipping | All |
| **weight_function** | `"tricube"` | 7 options | Distance kernel | All |
| **robustness_method** | `"bisquare"` | 3 options | Outlier weighting | All |
| **scaling_method** | `"mad"` | 3 options | Scale estimate | All |
| **zero_weight_fallback** | `"use_local_mean"` | 3 | Zero-weight | All |
| **boundary_policy** | `"extend"` | 4 options | Edge padding | All |
| **auto_converge** | `NULL` | tolerance | Early stopping | All |
| **return_residuals** | `FALSE` | logical | Include residuals | All |
| **return_robustness_weights** | `FALSE` | logical | Include weights | All |
| **return_se** | `FALSE` | logical | Return standard errors | All |
| **return_diagnostics** | `FALSE` | logical | Metrics | Batch, Streaming |
| **custom_weights** | `NULL` | positive | Per-observation weights | Batch |
| **backend** | `"cpu"` | `"cpu"`, `"gpu"` | Compute backend | Batch |
| **parallel** | `FALSE` | `TRUE`/`FALSE` | Parallel CPU execution | Batch |
| **confidence_intervals** | `NULL` | (0, 1) | CI coverage level | Batch |
| **prediction_intervals** | `NULL` | (0, 1) | PI level | Batch |
| **cv_method** | `NULL` | method | Auto-select fraction | Batch |
| **chunk_size** | 5000 | [10, ∞) | Points per chunk | Streaming |
| **overlap** | 500 | [0, chunk) | Overlap between chunks | Streaming |
| **merge_strategy** | `"weighted_average"` | 4 | Chunk merge | Streaming |
| **window_capacity** | 1000 | [3, ∞) | Max window size | Online |
| **min_points** | 2 | [2, window] | Min before output | Online |
| **update_mode** | `"incremental"` | 2 options | Update strategy | Online |

---

## Parameter Options

- **weight_function**: `"tricube"` (default), `"epanechnikov"`, `"gaussian"`,
  `"biweight"`, `"cosine"`, `"triangle"`, `"uniform"`
- **robustness_method**: `"bisquare"` (default), `"huber"`, `"talwar"`
- **zero_weight_fallback**: `"use_local_mean"` (default),
  `"return_original"`, `"return_none"`
- **boundary_policy**: `"extend"` (default), `"reflect"`, `"zero"`,
  `"noboundary"`
- **scaling_method**: `"mad"` (default), `"mar"`, `"mean"`
- **merge_strategy**: `"weighted_average"` (default), `"average"`,
  `"take_first"`, `"take_last"`
- **update_mode**: `"incremental"` (default), `"full"`

---

## Core Parameters

### fraction

The proportion of data used for each local fit. **Most important parameter.**

| Value | Effect | Use Case |
| --- | --- | --- |
| 0.1–0.3 | Fine detail | Rapidly changing signals |
| 0.3–0.5 | Balanced | General purpose |
| 0.5–0.7 | Heavy smoothing | Noisy data |
| 0.7–1.0 | Very smooth | Trend extraction |

```{r parameters_1}
library(rfastlowess)
set.seed(42)
x <- seq(0, 2 * pi, length.out = 100)
y <- sin(x) + rnorm(100, sd = 0.3)

model <- Lowess(fraction = 0.3)
result <- fit(model, x, y)
cat("First 6 smoothed values (fraction=0.3):\n")
print(head(result$y))
```

### iterations

Number of robustness iterations. Each iteration downweights high-residual
points.

```{r parameters_2}
model <- Lowess(iterations = 3)
result <- fit(model, x, y)
cat("First 6 smoothed values (iterations=3):\n")
print(head(result$y))
```

### delta

Distance threshold for skipping intermediate points (approximation). Setting
`delta > 0` speeds up computation on dense grids. Passed as a fraction of the
data range.

```{r parameters_3}
model <- Lowess(delta = 0.01)
result <- fit(model, x, y)
cat("First 6 smoothed values (delta=0.01):\n")
print(head(result$y))
```

### parallel

Enable parallel CPU execution (multiple cores).

```{r parameters_4}
model <- Lowess(parallel = TRUE)
result <- fit(model, x, y)
cat("First 6 smoothed values (parallel=TRUE):\n")
print(head(result$y))
```

### return_diagnostics

Return goodness-of-fit diagnostics (R², residuals, etc.).

```{r parameters_5}
model <- Lowess(return_diagnostics = TRUE)
result <- fit(model, x, y)
cat("R²:", result$diagnostics$r_squared, "\n")
```

---

## Streaming Parameters

### chunk_size and overlap

```{r parameters_6}
model <- StreamingLowess(
    chunk_size = 5000,
    overlap = 500,
    merge_strategy = "weighted_average"
)
```

---

## Online Parameters

### window_capacity and min_points

```{r parameters_7}
model <- OnlineLowess(
    window_capacity = 50,
    min_points = 5,
    update_mode = "incremental"
)
```

```{r session-info}
sessionInfo()
```
