---
title: "Robustness"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robustness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## How Robustness Works

Standard LOWESS can be biased by outliers. Robustness iterations downweight
points with large residuals:

1. Fit initial LOWESS
2. Compute residuals
3. Assign robustness weights (large residuals → low weight)
4. Refit using combined distance × robustness weights
5. Repeat steps 2–4

---

## Robustness Methods

![Robustness method comparison](../man/figures/robust_method_comparison.svg)

![Effect of robustness iterations](../man/figures/robust_iter_comparison.svg)

### Bisquare (Default)

Smooth downweighting. Points transition gradually from full weight to zero.

$$w(u) = \begin{cases} (1 - u^2)^2 & |u| < 1 \\ 0 & |u| \geq 1 \end{cases}$$

**Use when**: General purpose, balanced approach.

```{r robustness_1}
library(rfastlowess)
set.seed(42)
x <- seq(0, 2 * pi, length.out = 100)
y <- sin(x) + rnorm(100, sd = 0.3)

model <- Lowess(iterations = 3, robustness_method = "bisquare")
result <- fit(model, x, y)
cat("First 6 smoothed values (bisquare robustness):\n")
print(head(result$y))
```

### Huber

Less aggressive than bisquare; still downweights outliers but keeps moderate
residuals at reduced weight.

**Use when**: Mild outlier contamination; want to preserve moderate deviations.

```{r robustness_2}
model <- Lowess(iterations = 3, robustness_method = "huber")
result <- fit(model, x, y)
cat("First 6 smoothed values (huber robustness):\n")
print(head(result$y))
```

### Talwar

Hard thresholding — points above the threshold are excluded completely.

**Use when**: Known contamination that should be fully excluded.

```{r robustness_3}
model <- Lowess(iterations = 3, robustness_method = "talwar")
result <- fit(model, x, y)
cat("First 6 smoothed values (talwar robustness):\n")
print(head(result$y))
```

---

## Choosing the Number of Iterations

| Iterations | Effect | When to Use |
| --- | --- | --- |
| 0 | No robustness (fastest) | Clean data, speed-critical |
| 1–3 | Moderate | Most applications |
| 4–6 | Strong | Data with clear outliers |
| 7+ | Very strong | Heavy contamination |

```{r robustness_4}
library(rfastlowess)
set.seed(42)
x <- 1:100
y <- sin(x / 10) + rnorm(100, sd = 0.3)

# Inject outliers
y[c(25, 50, 75)] <- y[c(25, 50, 75)] + 5

# Without robustness
model_0 <- Lowess(iterations = 0)
result_0 <- fit(model_0, x, y)

# With robustness
model_3 <- Lowess(iterations = 3)
result_3 <- fit(model_3, x, y)

plot(x, y, pch = 16, col = "gray", main = "Effect of Robustness Iterations")
lines(result_0$x, result_0$y, col = "red", lwd = 2, lty = 2)
lines(result_3$x, result_3$y, col = "blue", lwd = 2)
legend("topright", c("Data", "iterations=0", "iterations=3"),
        pch = c(16, NA, NA), lty = c(NA, 2, 1),
        col = c("gray", "red", "blue"))
```

---

## Method Comparison

| Method | Handling | Best For |
| --- | --- | --- |
| `"bisquare"` | Smooth to zero | General purpose |
| `"huber"` | Linear then downweights | Mild contamination |
| `"talwar"` | Hard threshold (0/1) | Severe point outliers |

---

## Detecting Outliers

Use robustness weights to identify potential outliers:

```{r robustness_5}
library(rfastlowess)
set.seed(42)
x <- seq(0, 2 * pi, length.out = 100)
y <- sin(x) + rnorm(100, sd = 0.3)
y[c(20, 50, 80)] <- y[c(20, 50, 80)] + 5  # inject outliers

model <- Lowess(iterations = 5, return_robustness_weights = TRUE)
result <- fit(model, x, y)

for (i in seq_along(result$robustness_weights)) {
    if (result$robustness_weights[i] < 0.05)
        cat(sprintf("Point %d is likely an outlier (weight: %.3f)\n",
                    i, result$robustness_weights[i]))
}
```

```{r session-info}
sessionInfo()
```
