---
title: "Streaming Mode (StreamingLowess)"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Streaming Mode (StreamingLowess)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **See also:** [Choosing an Execution Mode](adapter-choice.html) for a comparison of all three modes.

Processes data in fixed-size chunks with configurable overlap. Results for each chunk are returned after calling `process_chunk()`. Call `finalize()` after the last chunk to flush remaining buffered points.

## When to Use

- Dataset is too large to fit in memory
- Processing data from a file or stream
- Batch pipeline with memory constraints

## Parameters

| Parameter | Default | Description |
| --- | --- | --- |
| `chunk_size` | 5000 | Points per chunk |
| `overlap` | 500 | Overlap between chunks |
| `merge_strategy` | `"weighted_average"` | How to merge overlapping values |

## Merge Strategies

| Strategy | Behavior |
| --- | --- |
| `"average"` | Arithmetic mean of both estimates |
| `"weighted_average"` | Distance-weighted blend (recommended) |
| `"take_first"` | Keep left-chunk estimate |
| `"take_last"` | Keep right-chunk estimate |

![Merge strategy comparison](../man/figures/merge_comparison.svg)

## Example

```{r streaming_1}
library(rfastlowess)
set.seed(42)
x <- seq(0, 2 * pi, length.out = 100)
y <- sin(x) + rnorm(100, sd = 0.3)
x_chunk <- x[seq_len(50)]
y_chunk <- y[seq_len(50)]

model <- StreamingLowess(
    fraction = 0.3,
    iterations = 2,
    chunk_size = 5000,
    overlap = 500,
    merge_strategy = "average"
)
result <- process_chunk(model, x_chunk, y_chunk)
final  <- finalize(model)
```

```{r session-info}
sessionInfo()
```
