---
title: "Use Case: Time Series Analysis"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Use Case: Time Series Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

LOWESS provides flexible trend extraction from time series without parametric assumptions. It handles irregular sampling, noise, and outliers naturally.

---

## Basic Trend Extraction

`fraction = 0.1` sizes the neighbourhood to 10% of the data at each evaluation point — narrow enough to follow a slowly varying trend without smearing periodic variation. Three robustness `iterations` down-weight noise spikes.

```{r use_case_time_series_1}
library(rfastlowess)

set.seed(42)
t <- seq(0, 100, length.out = 500)
trend <- 10 + 0.5 * t + 3 * sin(t / 10)
noise <- rnorm(500, sd = 3)
y <- trend + noise

model <- Lowess(fraction = 0.1, iterations = 3)
result <- fit(model, t, y)

plot(t, y, col = "gray", pch = ".",
     xlab = "Time", ylab = "Value", main = "Trend Extraction")
lines(result$x, result$y, col = "blue", lwd = 2)
legend("topleft", c("Observed", "Trend (LOWESS)"),
       pch = c(1, NA), lty = c(NA, 1), col = c("gray", "blue"))
```

---

## Seasonal Decomposition

Extract the trend component and compute the residual to inspect seasonality.

```{r use_case_time_series_2}
library(rfastlowess)
set.seed(42)

# Simulate monthly data with trend + seasonality + noise
t <- 1:120  # 10 years monthly
trend <- 100 + 0.5 * t
seasonal <- 15 * sin(2 * pi * t / 12)
noise <- rnorm(120, sd = 5)
y <- trend + seasonal + noise

# Extract trend with large fraction (heavy smoothing)
model <- Lowess(fraction = 0.4, iterations = 2)
result <- fit(model, t, y)

# Residual = observed - trend (should show seasonality)
residual <- y - result$y

par(mfrow = c(2, 1))
plot(t, y, type = "l", col = "gray",
     main = "Original + LOWESS Trend", xlab = "Month", ylab = "Value")
lines(result$x, result$y, col = "blue", lwd = 2)

plot(t, residual, type = "l", col = "darkgreen",
     main = "Detrended Residual (Seasonal Component)",
     xlab = "Month", ylab = "Residual")
abline(h = 0, lty = 2, col = "red")
par(mfrow = c(1, 1))
```

---

## Irregular Time Grids

LOWESS handles irregularly sampled data naturally — no interpolation needed.

```{r use_case_time_series_3}
library(rfastlowess)
set.seed(42)

# Irregular sampling: dense early, sparse later
t_irregular <- c(sort(runif(200, 0, 50)), sort(runif(50, 50, 100)))
y_irregular <- sin(t_irregular / 10) + rnorm(length(t_irregular), sd = 0.5)

model <- Lowess(fraction = 0.2, iterations = 2)
result <- fit(model, t_irregular, y_irregular)

plot(t_irregular, y_irregular, pch = 16, cex = 0.4, col = "gray",
     xlab = "Time", ylab = "Value", main = "Irregularly Sampled Time Series")
lines(result$x, result$y, col = "blue", lwd = 2)
```

---

## Uncertainty Bands for Forecasting Context

Prediction intervals widen the uncertainty band to include both the uncertainty in the fitted curve and the expected scatter of new observations. `fraction = 0.2` offers a balance between local detail and stable interval width.

```{r use_case_time_series_4}
library(rfastlowess)
set.seed(42)
t <- seq(0, 100, length.out = 500)
y <- 10 + 0.3 * t + sin(t / 5) + rnorm(500, sd = 2)

model <- Lowess(
    fraction = 0.2,
    iterations = 3,
    confidence_intervals = 0.95,
    prediction_intervals = 0.95
)
result <- fit(model, t, y)

plot(t, y, pch = ".", col = "gray",
     xlab = "Time", ylab = "Value",
     main = "Trend with 95% Confidence and Prediction Intervals")
polygon(c(result$x, rev(result$x)),
        c(result$prediction_upper, rev(result$prediction_lower)),
        col = rgb(0, 0, 1, 0.08), border = NA)
polygon(c(result$x, rev(result$x)),
        c(result$confidence_upper, rev(result$confidence_lower)),
        col = rgb(0, 0, 1, 0.20), border = NA)
lines(result$x, result$y, col = "blue", lwd = 2)
legend("topleft", c("PI 95%", "CI 95%", "Trend"),
       fill = c(rgb(0, 0, 1, 0.08), rgb(0, 0, 1, 0.20), NA),
       lty  = c(NA, NA, 1), col = c(NA, NA, "blue"), border = NA)
```

```{r session-info}
sessionInfo()
```
