---
title: "Use Case: Genomic Data Smoothing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use Case: Genomic Data Smoothing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- markdownlint-disable MD024 -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

LOWESS for methylation profiles, ChIP-seq signals, and other genomic data.

## Overview

Genomic data often contains noise from sequencing depth variation, PCR
artifacts, or biological heterogeneity. LOWESS smoothing helps reveal
underlying patterns.

---

## Methylation Profile Smoothing

### The Challenge

DNA methylation data (from bisulfite sequencing or arrays) shows
position-dependent patterns that can be obscured by measurement noise.

### Solution

A small `fraction = 0.1` lets LOWESS follow fine-scale spatial structure
without smearing the transitions between methylated and unmethylated regions.
`confidence_intervals = 0.95` produces uncertainty bands that naturally widen
at positions with sparser CpG coverage, making low-confidence segments
immediately apparent in the plot.

```{r use_case_genomics_1}
library(rfastlowess)

positions <- seq(0, 99000, by = 1000)
observed <- 50 + sin(positions / 1000) * 20 + 5

# positions and observed are your methylation data
model <- Lowess(
    fraction = 0.1,
    iterations = 3,
    confidence_intervals = 0.95
)
result <- fit(model, positions, observed)

# Smoothed profile in result$y
# CI bounds in result$confidence_lower/upper
plot(positions, observed, pch = ".", col = "gray",
    xlab = "Genomic Position (bp)", ylab = "Methylation Level",
    main = "Methylation Profile Smoothing")
lines(result$x, result$y, col = "blue", lwd = 2)
lines(result$x, result$confidence_lower, col = "blue", lty = 2)
lines(result$x, result$confidence_upper, col = "blue", lty = 2)
legend("topright", c("Observed", "Smoothed", "95% CI"),
        pch = c(1, NA, NA), lty = c(NA, 1, 2),
        col = c("gray", "blue", "blue"))
cat(sprintf(
    "95%% CI: [%.4f, %.4f]\n",
    result$confidence_lower[1], result$confidence_upper[1]
))
```

---

## ChIP-seq Signal Smoothing

### Application

ChIP-seq experiments produce sparse, noisy coverage data. LOWESS can help
identify binding regions.

`fraction = 0.05` provides high spatial resolution — important for resolving
narrow binding peaks that would otherwise be smeared into the background. The
larger `iterations = 5` is deliberate: Poisson-distributed read counts produce
tall, isolated spikes, and extra robustness iterations progressively down-weight
them so the estimated background level is not inflated by a handful of extreme
counts.

```{r use_case_genomics_2}
library(rfastlowess)

positions <- seq(0, 99000, by = 1000)
signal <- 50 + sin(positions / 1000) * 20 + 5

model <- Lowess(
    fraction = 0.05,
    iterations = 5
)
result <- fit(model, positions, signal)

# Identify peaks above threshold
peak_count <- sum(result$y > 65.0)
cat(sprintf("y[0]: %.4f\n", result$y[1]))
cat(sprintf("Peak count: %d\n", peak_count))
```

---

## Large Genome Coverage (Streaming)

For whole-genome data that doesn't fit in memory:

```{r use_case_genomics_3}
library(rfastlowess)

positions <- seq(0, 10000, by = 10)
coverage <- 50 + sin(positions / 100) * 20 + 5.0

model <- StreamingLowess(
    fraction = 0.05,
    iterations = 3,
    chunk_size = 50,
    overlap = 10,
    merge_strategy = "weighted_average"
)

process_chunk(model, positions, coverage)
result <- finalize(model)
cat(sprintf("y[0]: %.4f\n", result$y[1]))
```

---

## Best Practices for Genomic Data

| Consideration | Recommendation |
| --- | --- |
| **Fraction** | 0.05–0.15 (preserve local features) |
| **Iterations** | 3–5 (handle sequencing outliers) |
| **Large data** | Use streaming mode |
| **Sparse regions** | Use `boundary_policy="extend"` |
| **Multiple chromosomes** | Process separately or ensure sorted |

---

## See Also

- [Concepts](concepts.html) — How LOWESS works
- `?Lowess` — All options
- [Robustness](robustness.html) — Outlier downweighting in depth
- [Merge Strategies](merge.html) — Streaming chunk reconciliation
- [Boundary Handling](boundary.html) — Edge handling for sparse regions
- [Real-Time Processing](use-case-real-time.html) — For sequencing runs

```{r session-info}
sessionInfo()
```
