---
title: "Use Case: Genomic Data Smoothing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use Case: Genomic Data Smoothing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

LOWESS is well-suited for genomic data such as DNA methylation profiles,
ChIP-seq signals, and expression arrays. Sequencing depth variation, PCR
artifacts, and biological heterogeneity all create noise that LOWESS handles
gracefully without parametric assumptions.

---

## Methylation Profile Smoothing

A small `fraction = 0.1` lets LOWESS follow fine-scale spatial structure without
smearing transitions between methylated and unmethylated regions.
`confidence_intervals = 0.95` produces uncertainty bands that naturally widen at
positions with sparser CpG coverage.

```{r use_case_genomics_1}
library(rfastlowess)

# Deterministic methylation profile along a chromosome
positions <- seq(0, 99000, by = 1000)
observed <- (50 + sin(positions / 1000) * 20 + 5) / 100  # Methylation is 0-1

# Smooth
model <- Lowess(
    fraction = 0.1,
    iterations = 3,
    confidence_intervals = 0.95
)
result <- fit(model, positions, observed)

# Plot
plot(positions, observed, pch = ".", col = "gray",
    xlab = "Genomic Position (bp)", ylab = "Methylation Level",
    main = "Methylation Profile Smoothing")
lines(result$x, result$y, col = "blue", lwd = 2)
lines(result$x, result$confidence_lower, col = "blue", lty = 2)
lines(result$x, result$confidence_upper, col = "blue", lty = 2)
legend("topright", c("Observed", "Smoothed", "95% CI"),
        pch = c(1, NA, NA), lty = c(NA, 1, 2),
        col = c("gray", "blue", "blue"))
cat(sprintf("95%% CI: [%.4f, %.4f]\n", result$confidence_lower[1], result$confidence_upper[1]))
```

---

## ChIP-seq Signal Smoothing

ChIP-seq experiments produce sparse, noisy coverage data. LOWESS can help
identify binding regions.

`fraction = 0.05` provides high spatial resolution — important for resolving
narrow binding peaks that would otherwise be smeared into the background. The
larger `iterations = 5` is deliberate: Poisson-distributed read counts produce
tall, isolated spikes, and extra robustness iterations progressively down-weight
them so the estimated background level is not inflated by a handful of extreme
counts.

```{r use_case_genomics_2}
library(rfastlowess)

# Deterministic ChIP-seq-style coverage profile
positions <- seq(0, 99000, by = 1000)
signal <- 50 + sin(positions / 1000) * 20 + 5

# Smooth
model <- Lowess(
    fraction = 0.05,
    iterations = 5
)
result <- fit(model, positions, signal)
peak_count <- sum(result$y > 65.0)

plot(positions, signal, pch = 16, cex = 0.4, col = "gray",
    xlab = "Genomic Position", ylab = "Read Count",
    main = "ChIP-seq Signal Smoothing")
lines(result$x, result$y, col = "blue", lwd = 2)
abline(h = 65.0, col = "red", lty = 2)
cat(sprintf("y[0]: %.4f\n", result$y[1]))
cat(sprintf("Peak count: %d\n", peak_count))
```

---

## Large Genome Coverage (Streaming)

For whole-genome datasets that do not fit in memory, use the streaming adapter.

```{r use_case_genomics_3}
library(rfastlowess)

model <- StreamingLowess(
    fraction   = 0.05,
    iterations = 3,
    chunk_size = 50,
    overlap    = 10,
    merge_strategy = "weighted_average"
)

# Deterministic whole-genome coverage, processed as a single chunk
positions <- seq(0, 10000, by = 10)
coverage <- 50 + sin(positions / 100) * 20 + 5
result <- process_chunk(model, positions, coverage)
final <- finalize(model)
cat(sprintf("y[0]: %.4f\n", final$y[1]))
```

```{r session-info}
sessionInfo()
```
