---
title: "Use Case: Genomic Data Smoothing"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Use Case: Genomic Data Smoothing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

LOWESS is well-suited for genomic data such as DNA methylation profiles,
ChIP-seq signals, and expression arrays. Sequencing depth variation, PCR
artifacts, and biological heterogeneity all create noise that LOWESS handles
gracefully without parametric assumptions.

---

## Methylation Profile Smoothing

A small `fraction = 0.1` lets LOWESS follow fine-scale spatial structure without
smearing transitions between methylated and unmethylated regions.
`confidence_intervals = 0.95` produces uncertainty bands that naturally widen at
positions with sparser CpG coverage.

```{r use_case_genomics_1}
library(rfastlowess)

# Simulate methylation data along a chromosome
set.seed(42)
n <- 1000
positions <- sort(runif(n, 0, 1e6))

# True methylation pattern
true_meth <- 0.5 + 0.3 * sin(positions / 1e5)

# Observed with noise
observed <- true_meth + rnorm(n, sd = 0.15)
observed <- pmax(0, pmin(1, observed))

# Smooth
model <- Lowess(
    fraction = 0.1,
    iterations = 3,
    confidence_intervals = 0.95
)
result <- fit(model, positions, observed)

# Plot
plot(positions, observed, pch = ".", col = "gray",
    xlab = "Genomic Position (bp)", ylab = "Methylation Level",
    main = "Methylation Profile Smoothing")
lines(result$x, result$y, col = "blue", lwd = 2)
lines(result$x, result$confidence_lower, col = "blue", lty = 2)
lines(result$x, result$confidence_upper, col = "blue", lty = 2)
legend("topright", c("Observed", "Smoothed", "95% CI"),
        pch = c(1, NA, NA), lty = c(NA, 1, 2),
        col = c("gray", "blue", "blue"))
```

---

## ChIP-seq Signal Smoothing

ChIP-seq experiments produce sparse, noisy coverage data. LOWESS can help
identify binding regions.

`fraction = 0.05` provides high spatial resolution — important for resolving
narrow binding peaks that would otherwise be smeared into the background. The
larger `iterations = 5` is deliberate: Poisson-distributed read counts produce
tall, isolated spikes, and extra robustness iterations progressively down-weight
them so the estimated background level is not inflated by a handful of extreme
counts.

```{r use_case_genomics_2}
library(rfastlowess)
set.seed(42)

# Simulated ChIP-seq enrichment
n <- 500
positions <- seq(1, 50000, length.out = n)
signal <- 10 + 5 * dnorm(positions, mean = 25000, sd = 5000) * 1000

# Add noise and spikes
noise <- rpois(n, lambda = 2)
signal_noisy <- signal + noise
signal_noisy[c(100, 200, 300)] <- signal_noisy[c(100, 200, 300)] * 5

# Smooth
model <- Lowess(
    fraction = 0.05,
    iterations = 5
)
threshold <- quantile(result$y, 0.75)

plot(positions, signal_noisy, pch = 16, cex = 0.4, col = "gray",
    xlab = "Genomic Position", ylab = "Read Count",
    main = "ChIP-seq Signal Smoothing")
lines(result$x, result$y, col = "blue", lwd = 2)
abline(h = threshold, col = "red", lty = 2)
```

---

## Large Genome Coverage (Streaming)

For whole-genome datasets that do not fit in memory, use the streaming adapter.

```{r use_case_genomics_3}
library(rfastlowess)

model <- StreamingLowess(
    fraction   = 0.05,
    chunk_size = 100000,
    overlap    = 10000,
    merge_strategy = "weighted_average"
)

# Process chromosome in chunks
# (in practice, read from a file or database)
set.seed(42)
for (chunk_i in 1:5) {
    pos_chunk <- seq((chunk_i - 1) * 1e5 + 1, chunk_i * 1e5, length.out = 1000)
    val_chunk <- sin(pos_chunk / 1e4) + rnorm(1000, sd = 0.2)
    result <- process_chunk(model, pos_chunk, val_chunk)
}
final <- finalize(model)
print(head(final$y))
```

```{r session-info}
sessionInfo()
```
