---
title: "Use Case: Real-Time Processing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use Case: Real-Time Processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

When data arrives continuously from sensors, logs, or streaming pipelines, the
Online adapter provides incremental smoothing without reprocessing the entire
dataset.

---

## Online Mode: Point-by-Point

`window_capacity = 25` limits the internal buffer to the 25 most recent
observations; each `add_point()` call costs O(window) rather than growing with
total history. `min_points = 5` suppresses output until the window holds enough
points for a stable fit. `update_mode = "incremental"` re-fits only the most
recent point rather than the full window, halving typical latency.

### Sensor Data Example

```{r use_case_real_time_1}
library(rfastlowess)

times <- 0:99
temperatures <- 20 + 5 * sin(times / 10) + sin(times * 1.7) * 0.5

model <- OnlineLowess(
    fraction = 0.3,
    iterations = 1,
    window_capacity = 25,
    min_points = 5,
    update_mode = "incremental"
)
count <- 0
for (i in seq_along(times)) {
    result <- add_point(model, times[i], temperatures[i])
    if (!is.null(result)) {
        if (count < 5) {
            cat(sprintf("Time %d: smoothed = %.4f\n", times[i], result$y))
        }
        count <- count + 1
    }
}
cat(sprintf("... (%d more)\n", count - 5))
```

---

## Accumulating Results

Collect online smoothed values into a vector for downstream analysis or
plotting.

```{r use_case_real_time_2}
library(rfastlowess)
set.seed(42)
n <- 200
times <- seq_len(n)
signal <- 10 + 2 * sin(times / 20) + rnorm(n, sd = 1)

model <- OnlineLowess(
    fraction       = 0.3,
    window_capacity = 30,
    min_points      = 5
)

smoothed_x <- numeric(n)
smoothed_y <- numeric(n)
n_out <- 0L

for (i in seq_len(n)) {
    result <- add_point(model, times[i], signal[i])
    if (!is.null(result)) {
        n_out <- n_out + 1L
        smoothed_x[n_out] <- times[i]
        smoothed_y[n_out] <- result$y
    }
}

smoothed_x <- smoothed_x[seq_len(n_out)]
smoothed_y <- smoothed_y[seq_len(n_out)]

plot(times, signal, pch = 16, cex = 0.4, col = "gray",
    xlab = "Time", ylab = "Value",
    main = "Online LOWESS — Accumulated Output")
lines(smoothed_x, smoothed_y, col = "blue", lwd = 2)
```

---

## Update Modes

| Mode | Behaviour | Latency | Accuracy |
| --- | --- | --- | --- |
| `"incremental"` | Re-fits only the newest point | Low | Moderate |
| `"full"` | Re-fits the entire window | Higher | Higher |

```{r use_case_real_time_3}
# Full-window refit (more accurate, slower)
model_full <- OnlineLowess(
    fraction       = 0.3,
    window_capacity = 25,
    update_mode     = "full"
)

smoothed_full_x <- numeric(n)
smoothed_full_y <- numeric(n)
n_full <- 0L

for (i in seq_len(n)) {
    result <- add_point(model_full, times[i], signal[i])
    if (!is.null(result)) {
        n_full <- n_full + 1L
        smoothed_full_x[n_full] <- times[i]
        smoothed_full_y[n_full] <- result$y
    }
}

smoothed_full_x <- smoothed_full_x[seq_len(n_full)]
smoothed_full_y <- smoothed_full_y[seq_len(n_full)]

plot(times, signal, pch = 16, cex = 0.4, col = "gray",
    xlab = "Time", ylab = "Value",
    main = "Online LOWESS — Full-Window Refit")
lines(smoothed_full_x, smoothed_full_y, col = "blue", lwd = 2)
```

---

## Streaming Mode for Large Batches

> **Note:** Always call `finalize()` after the last chunk to flush any remaining
> buffered points and produce output for the tail of the dataset.

For large historical files arriving in chunks (not true real-time), the
Streaming adapter is more efficient than Online:

```{r use_case_real_time_4}
library(rfastlowess)

model <- StreamingLowess(
    fraction       = 0.1,
    iterations     = 2,
    chunk_size     = 50,
    overlap        = 10,
    merge_strategy = "weighted_average"
)

chunk1_x <- 0:49
chunk1_y <- sin(chunk1_x) + 0.1
chunk2_x <- 50:99
chunk2_y <- sin(chunk2_x) + 0.1

process_chunk(model, chunk1_x, chunk1_y)
process_chunk(model, chunk2_x, chunk2_y)

final <- finalize(model)
cat(sprintf("y[0]: %.6f\n", final$y[1]))
```

```{r session-info}
sessionInfo()
```
