---
title: "Use Case: Time Series Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use Case: Time Series Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- markdownlint-disable MD024 -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

LOWESS for trend extraction and temporal smoothing.

## Overview

Time series data often contains noise, seasonality, and trends. LOWESS
provides flexible trend extraction without parametric assumptions.

---

## Basic Trend Extraction

`fraction = 0.1` sizes the neighbourhood as 10% of the data at each evaluation
point — narrow enough to follow a slowly varying trend without smearing
periodic variation. Three robustness `iterations` down-weight noise spikes so
they cannot bias the fitted curve; this is especially important when the
signal-to-noise ratio is low or when occasional outliers are expected.

```{r use_case_time_series_1}
library(rfastlowess)

n <- 500
t <- seq(0, 100, length.out = n)
i <- 0:(n - 1)
y <- 10 + 0.5 * t + 3 * sin(t / 10) + ((i * 7 + 3) %% 1.7 - 0.85) * 3

# t and y are your time series vectors
model <- Lowess(
    fraction = 0.1,
    iterations = 3
)
result <- fit(model, t, y)

cat(sprintf("y[0]: %.4f\n", result$y[1]))
```

---

## Detrending

Remove trend to analyze residual patterns.

Setting `return_residuals = TRUE` stores `observed - smoothed` alongside the
smooth. A slightly wider `fraction = 0.3` produces a smoother baseline trend, so
short-duration oscillations end up in the residuals rather than being absorbed
into the trend component. The residual series is then ready for spectral
analysis, seasonality detection, or change-point methods.

```{r use_case_time_series_2}
library(rfastlowess)

n <- 100
t <- seq(0, 2 * pi, length.out = n)
y <- sin(t) + 0.1

model <- Lowess(
    fraction = 0.3,
    iterations = 3,
    return_residuals = TRUE
)
result <- fit(model, t, y)

trend <- result$y
detrended <- result$residuals
cat(sprintf("Trend y[0]: %.4f  residual: %.4f\n", trend[1], detrended[1]))
```

---

## Forecasting with Prediction Intervals

Prediction intervals widen the uncertainty band to include both the
uncertainty in the fitted curve (confidence interval) and the expected scatter
of new observations around it. `fraction = 0.2` offers a balance between local
detail and stable interval width — too small a fraction produces jagged
interval edges; too large a fraction underestimates local variance near turning
points.

```{r use_case_time_series_3}
library(rfastlowess)

n <- 100
t <- seq(0, 2 * pi, length.out = n)
y <- sin(t) + 0.1

model <- Lowess(
    fraction = 0.2,
    iterations = 3,
    confidence_intervals = 0.95,
    prediction_intervals = 0.95
)
result <- fit(model, t, y)

cat(sprintf(
    "95%% PI: [%.4f, %.4f]\n",
    result$prediction_lower[1], result$prediction_upper[1]
))
```

---

## Handling Missing Data

LOWESS naturally handles irregular time sampling:

```{r use_case_time_series_4}
library(rfastlowess)

t_irregular <- vapply(
    0:99, function(i) i * 1.0 + (i * 31 %% 10) * 0.1, numeric(1)
)
y_irregular <- 10 + 0.3 * t_irregular + 2.0 * sin(t_irregular * 0.1)

# No special handling needed for irregular spacing
model <- Lowess(fraction = 0.2)
result <- fit(model, t_irregular, y_irregular)
cat(sprintf("y[0]: %.4f\n", result$y[1]))
```

---

## Multi-Scale Analysis

Use different fractions to extract features at different scales:

```{r use_case_time_series_5}
library(rfastlowess)

n <- 100
t <- seq(0, 2 * pi, length.out = n)
y <- sin(t) + 0.1

scales <- c(0.05, 0.2, 0.5)
trends <- lapply(scales, function(f) {
    model <- Lowess(fraction = f)
    fit(model, t, y)$y
})
cat("Trend y[0] values:", sapply(trends, function(tr) tr[1]), "\n")
```

---

## Gene Expression Time Course

Biological application:

```{r use_case_time_series_6}
library(rfastlowess)

hours <- seq(0, 24, by = 0.5)
i <- seq_along(hours) - 1
expression <- 100 * (1 + 0.5 * sin(hours * pi / 12)) +
    ((i * 7 + 3) %% 1.7 - 0.85) * 10

model <- Lowess(
    fraction = 0.3,
    iterations = 3,
    confidence_intervals = 0.95,
    return_diagnostics = TRUE
)
result <- fit(model, hours, expression)
cat(sprintf("R2: %.3f\n", result$diagnostics$r_squared))
```

---

## Choosing Fraction for Time Series

| Data Type | Recommended Fraction | Rationale |
| --- | --- | --- |
| Daily data (years) | 0.3–0.5 | Capture annual trends |
| Hourly data (days) | 0.1–0.2 | Capture daily patterns |
| Sensor data (minutes) | 0.05–0.1 | Preserve short-term features |
| Noisy data | Higher | Reduce noise impact |
| Clean data | Lower | Preserve detail |

---

## See Also

- [Real-Time Processing](use-case-real-time.html) — For streaming time series
- [Cross-Validation](cross-validation.html) — Optimal fraction selection
- [Boundary Handling](boundary.html) — Edge bias in trend extraction
- `?Lowess` — Full parameter reference

```{r session-info}
sessionInfo()
```
